Longitudinal Manifold Estimation

### 2023-04-17 paper/main.tex
add lpme initialization explanation

To solve this problem, we propose an extension of the method developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients for spline functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Details of this approach are described below.

\textit{Initialization}:

>>>>

To solve this problem, we propose an extension of the method developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Details of this approach are described below.

\textit{Initialization}: We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$. Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers from $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$. HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure. The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

The PME algorithm uses ISOMAP (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed. Instead of using ISOMAP on the data for each time point separately, the LMPE approach removes the time dimension from the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously. Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.
